Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-11-12 paper-bedpca.tex
save

\title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
\author{Florian Priv\'e,$^{\text{1,}*}$ Keurcien Luu, Michael G.B. Blum,$^{\text{4}}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}

@@

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

@@

For example, we show that PC19 to PC40 in the UK Biobank are capturing LD structure. Using our automatic algorithm for removing long-range LD region, we are able to recover only 16 PCs that capture population structure only.
Therefore, we recommend using only 16-18 PCs from the UK Biobank. Another example: when projecting the PCA computed on the 1000 Genomes project, although PC1 to PC4 suffer from only moderate shrinkage (1.00-1.08), PC5 for example suffers from a shrinkage factor of 1.51 and PC10 from a shrinkage factor of 3.16. We provide a fast way to project new individuals and that is not affected by this shrinkage bias. We also show how to restrict to individuals that are genetically homogeneous based on PCA.

@@

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates has been used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software address all of these. 
In the following, we recall common pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}. Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.  
Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics. We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on bed/bim/fam PLINK files with some missing values. 
We use them to analyse the UK Biobank data, and show that these functions are both very fast and easy to use. We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure. Interestingly, subsetting the UK Biobank data enables to get more PCs capturing population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to get unbiased projections instead.
Finally, we talk about detecting outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to homogeneous individuals.

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods.} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|}
\hline
Analysis & Method and/or Software & Advantages & Limitations \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
 & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT \cite[]{patterson2006population}} & Simple & Assume Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
  & Bias-adjusted projection in R package hdpca \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition and Procrustes transformation \\ (ADP) in LASER 2.0 \cite[]{wang2015improved}} & Accurate & Slow (a new PCA for each new sample) \\
  & Online ADP (OADP) in python package FRAPOSA \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$). Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).

When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.

To compute this decomposition easily and efficiently, we implement an accessor that memory-map the bed file to use it directly as if it were a standard matrix. Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}. This method is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotyped matrix with a given vector, which we implement for bed files and parallelise.

@@

Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance of $x$.
We use these distances for many applications in this paper. 
When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
If $x$ are PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}. We reexport function \texttt{covRob} in R package bigutilsr for convenience.

@@

Finally, to choose the threshold of outlierness based on the previously described statistics, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}. 
Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster. 
In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
Automatic outlier detection is difficult, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).

@@

To demonstrate that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307. It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
When performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).

@@

To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}).
6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
[WHERE TO PUT FOLLOWING?] In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

@@

We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs). Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}). 
When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is an apparent bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP. Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).

@@

When computing the PCA on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}). Shrinkage increases with an increased number of variants used, but decreases with an increased sample size (Table \ref{tab:shrinkage}).

\begingroup
\renewcommand*{\arraystretch}{1.8}
\begin{table}[h]
\centering
\caption{Shrinkage coefficients when projecting new individuals onto reference PCA space. We list the dataset, the sample size and number of variants used to compute the final truncated PCA.} 
\label{tab:shrinkage}
\begin{tabular}{|c|c|c|c|}
\hline
Dataset & Sample size (x1000) & \makecell {Number of variants \\ (x1000, after LD removal)} & Shrinkage (PC 1 - 5 - 10 - 20) \\
\hline
1000G & 1.5 & 393 & 1.01 - 1.50 - 3.14 - 6.70 \\
1000G & 2.5 & 229 & 1.01 - 1.36 - 2.84 - 6.75 \\
UKBB & 49 & 282 & 1.00 - 1.04 - 1.12 - 1.43 \\
UKBB & 406.5 & 172 & 1.00 - 1.01 - 1.04 - 1.08 \\
\hline
\end{tabular}
\end{table}
\endgroup

@@

We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unkonwn self-reported ancestry. 
We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
We show that we are able to capture more PCs (at least 40) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection. Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB. Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).

@@

>>>>

\title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
\author{Florian Priv\'e,$^{\text{1,5,}*}$ Keurcien Luu$^{\text{5}}$, Michael G.B. Blum,$^{\text{4,5}}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}

@@

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche, 38700, France. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

@@

For example, we show that PC19 to PC40 in the UK Biobank capture LD structure. Using our automatic algorithm for removing long-range LD region, we recover 16 PCs that capture population structure only.
Therefore, we recommend using only 16-18 PCs from the UK Biobank. 
We provide evidence for a shrinkage bias when projecting PCs computed with data from the 1000 Genomes project.  Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 for example suffers from a shrinkage factor of 1.50 and PC10 from a shrinkage factor of 3.14. 
We provide a fast way to project new individuals and that is not affected by this shrinkage bias. 
We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.

@@

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software address all of these. 
In the following, we outline common pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}. Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.  
Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics. We make these available in a new release of R package bigsnpr (v1.0.0); this package differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on bed/bim/fam PLINK files with some missing values. 
We use them to analyse the UK Biobank data, and show that these functions are both very fast and easy to use. We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure. Interestingly, subsetting the UK Biobank data enables to get more PCs capturing population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to estimate unbiased projections instead.
Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to individuals of homogeneous ancestry.

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods.} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|c|}
\hline
Analysis & Method and/or Software & Citation & Advantages & Limitations \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
 & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assume Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix of diploid individuals $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$). Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).

When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non-missing values of each variant.
Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.

To compute this decomposition easily and efficiently, we implement an accessor that memory-map the PLINK bed file to use it directly as if it were a standard matrix. Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}. This method is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotyped matrix with a given vector, which we implement for PLINK bed files and parallelise.

@@

Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance of $x$.
We use these distances for many applications in this paper. 
When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
If $x$ are PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters, as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}. 
We reexport function \texttt{covRob} in R package bigutilsr for convenience.

@@

Finally, to choose the threshold on the previously described statistics above which to be considered as outlier, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}. 
Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster. 
In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics).

@@

To demonstrate that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations (CPU: Intel(R) Xeon(R) Silver 4114, 2.20GHz).
It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307. It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
When performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).

@@

To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}).
6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic used to define outliers.\\~\\}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

@@

We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCs ($K=20$ PCs). Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}). 
When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is an apparent bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP. Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).

@@

When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}). Shrinkage increases with an increased number of variants used, but decreases with an increased sample size (Table \ref{tab:shrinkage}).

\begingroup
\renewcommand*{\arraystretch}{1.8}
\begin{table}[h]
\centering
\caption{Shrinkage coefficients when projecting new individuals onto reference PCA space. We list the dataset, the sample size and number of variants used to compute the final truncated PCA. As expected, the shrinkage bias only becomes negligible if the PCA is conducted on large samples.} 
\label{tab:shrinkage}
\begin{tabular}{|c|c|c|c|}
\hline
Dataset & Sample size (x1000) & \makecell {Number of variants \\ (x1000, after LD removal)} & Shrinkage (PC 1 - 5 - 10 - 20) \\
\hline
1000G & 1.5 & 393 & 1.01 - 1.50 - 3.14 - 6.70 \\
1000G & 2.5 & 229 & 1.01 - 1.36 - 2.84 - 6.75 \\
UKBB & 49 & 282 & 1.00 - 1.04 - 1.12 - 1.43 \\
UKBB & 406.5 & 172 & 1.00 - 1.01 - 1.04 - 1.08 \\
\hline
\end{tabular}
\end{table}
\endgroup

@@

We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unknown self-reported ancestry. 
We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
We show that we are able to capture more PCs (at least 40) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection. Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB. Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).

@@

\noindent[TODO: INCLUDE CITATIONS OF EVEN SAMPLING IN PCA]
